Verification of OpenSSL's SHA3 Keccak function using the Software Analysis Workbench

### 2021-09-16 paper/fips.tex
Restructure sections
Fix more LaTeX apostrophe issues

\section{Amended Cryptol}
\label{sec:amended}
Initially it appeared that visual inspection with Cryptol’s standard syntax was subpar.
The language's functional folds and list comprehensions do not align easily with the imperative 
and array-based NIST standard. By leveraging Cryptol’s versatility, however, we produced a couple 
utility functions to visually mimic the sequential nature of the NIST specification.

\subsection{For Utility Function}
The phrase “for all” indicates a looping structure and is quite common in the NIST standard. 
It is used in the definition of the pi function shown in Figure \ref{fig:nistPi}.

@@

Cryptol declares list comprehension variables after they are used, and the order of the
nested loops is not as apparent. Using a helper function, ‘for’, we were able to provide
a more clear definition seen in Figure \ref{fig:cryptolAmendedPi}.

@@

The ‘for’ utility function follows the same intuition as the implementation in most 
imperative languages in that it iterates and runs a loop body over defined elements. 
It recieves a function and a list of arguments and returns a list of results from calling the function 
on every element of the input array. Figure \ref{fig:cryptolFor} shows it's implementation.

@@

\subsection{While Utility Function}
Another common phrasing is the “let...for...let”. This defines a state upon which a folding 
occurs, iteratively condensing the information into a final state. Imperatively, it is most 
comparable to a while loop. As shown by Figure \ref{fig:nistRC}, is present in the rc function 
defined by the NIST specification.

@@

Maintaining this state through Cryptol's list comprehensions can be difficult.
The comprehensions require a list to iterate through.Since our ‘for’ utility 
function is simply a wrapper for list comprehensions, it is equally unsuited for 
visual verification on such steps. The condition to stop iterating in the RC function
bounds the number of iterations by [1, 254], which lets us construct such a list to 
iterate through. This leads to a rather ungainly Cryptol definition, however, as 
shown by Figure \ref{fig:cryptolRC}.

@@

Not only is this definition computationally slower as it runs 255 loops every time it is called, 
it also awkwardly indexes into a list of previous states. This differs vastly from the 
NIST specificiation both visually and functionally. Using a new ‘while’ utility function, we 
can declare an initial state, condition function, and body function that will recursively 
be called until the condition is met. It consolodates the information of the previous state 
until the condition is sastified. This definition is foudn in Figure \ref{fig:cryptolamendedRC}.

@@

The final state is returned by the ‘while’ function. This iterates only until the 
condition is met, avoids the clutter of indexing within a larger array of all previous 
states, and places the condition in a more obvious location before the loop body. This 
utility function is defined in Figure \ref{fig:cryptolWhile}.

>>>>

\section{SHA3 FIPS Specification in Cryptol}
\label{sec:fips}
Initially it appeared that visual inspection with Cryptol's standard syntax was subpar.
The language's functional folds and list comprehensions do not align easily with the imperative 
and array-based NIST standard. By leveraging Cryptol's versatility, however, we produced a couple 
utility functions to visually mimic the sequential nature of the NIST specification.

\subsection{For Utility Function}
The phrase \emph{for all} indicates a looping structure and is quite common in the NIST standard. 
It is used in the definition of the pi function shown in Figure \ref{fig:nistPi}.

@@

Cryptol declares list comprehension variables after they are used, and the order of the
nested loops is not as apparent. Using a helper function, \emph{for}, we were able to provide
a more clear definition seen in Figure \ref{fig:cryptolAmendedPi}.

@@

The \emph{for} utility function follows the same intuition as the implementation in most 
imperative languages in that it iterates and runs a loop body over defined elements. 
It recieves a function and a list of arguments and returns a list of results from calling the function 
on every element of the input array. Figure \ref{fig:cryptolFor} shows it's implementation.

@@

\subsection{While Utility Function}
Another common phrasing is the \emph{let...for...let}. This defines a state upon which a folding 
occurs, iteratively condensing the information into a final state. Imperatively, it is most 
comparable to a while loop. As shown by Figure \ref{fig:nistRC}, is present in the rc function 
defined by the NIST specification.

@@

Maintaining this state through Cryptol's list comprehensions can be difficult.
The comprehensions require a list to iterate through.Since our \emph{for} utility 
function is simply a wrapper for list comprehensions, it is equally unsuited for 
visual verification on such steps. The condition to stop iterating in the RC function
bounds the number of iterations by [1, 254], which lets us construct such a list to 
iterate through. This leads to a rather ungainly Cryptol definition, however, as 
shown by Figure \ref{fig:cryptolRC}.

@@

Not only is this definition computationally slower as it runs 255 loops every time it is called, 
it also awkwardly indexes into a list of previous states. This differs vastly from the 
NIST specificiation both visually and functionally. Using a new \emph{while} utility function, we 
can declare an initial state, condition function, and body function that will recursively 
be called until the condition is met. It consolodates the information of the previous state 
until the condition is sastified. This definition is foudn in Figure \ref{fig:cryptolamendedRC}.

@@

The final state is returned by the \emph{while} function. This iterates only until the 
condition is met, avoids the clutter of indexing within a larger array of all previous 
states, and places the condition in a more obvious location before the loop body. This 
utility function is defined in Figure \ref{fig:cryptolWhile}.

### 2021-09-16 paper/proof.tex
Fix more LaTeX apostrophe issues

The Software Analysis Workbench (SAW) provides a verification environment for compositional proofs of functional equivalence between two pieces of supported code (C, Java, or Cryptol).  
SAW uses symbolic execution to create computational models of the code and creates problems of functional equivalence between those two models that can be pushed off to an SMT solver.  
The difficulty of these SMT problems is directly affected by the size and complexity of the computations that are being compared.  
This can lead to drastic increases in proof times and sometimes even prevents completion as computations increase.  
SAW’s solution to this is to create contracts (called overrides) of a functions input to output relationship which allows for a modular approach to writing the proof.  
By breaking up computation into smaller pieces, and creating overrides for these pieces, SAW can often simplify an otherwise complex (and thus incompletable) proof to remain within the capacity of an SMT solver.

>>>>

SAW provides a verification environment for compositional proofs of functional equivalence between two pieces of supported code (C, Java, or Cryptol).  
SAW uses symbolic execution to create computational models of the code and creates problems of functional equivalence between those two models that can be pushed off to an SMT solver.  
The difficulty of these SMT problems is directly affected by the size and complexity of the computations that are being compared.  
This can lead to drastic increases in proof times and sometimes even prevents completion as computations increase.  
SAW's solution to this is to create contracts (called overrides) of a functions input to output relationship which allows for a modular approach to writing the proof.  
By breaking up computation into smaller pieces, and creating overrides for these pieces, SAW can often simplify an otherwise complex (and thus incompletable) proof to remain within the capacity of an SMT solver.

### 2021-09-16 paper/results.tex
Restructure sections
Fix more LaTeX apostrophe issues

* Visual inspection and limited test vectors for (1) and (2)
* Only provide proof certificates for message sizes of X to Y for digest sizes of … for (3)
* Visual inspection only for anything above Keccak for (3) in OpenSSLs C-code

Effort (see above and then consider below):
 - How long were we working on the Cryptol code?
 - How long were we working on the SAW proofs?
 - How long did it take to understand the OpenSSL code?
We'll want to determine how long it took us and then try to infer how long that would have taken had we been working full time (40 hours/week).

Because of the modular design of the SHA-3 specification, the Cryptol implementation and OpenSSL implementation are programmatically more similar. The only strong discrepancy is that the OpenSSL implemenation organizes the bits of the sponge construction. The affect of modular cryptographic design can be seen by the proof completing successfully in the Yices, Z3, and ABC SMT-solvers. Previous work on SHA2-256 was only successful with Yices \ref{cite:nfm-us}. Full results are show in table \ref{finaltable}.

\begin{table}[b]
\caption{Proof Runtimes for SHA-3 with SMT-solvers}\label{finaltable}
\setlength{\tabcolsep}{13.5pt}
\begin{tabular}{|l|r|r|r|}
\hline
\textbf{Function} & \textbf{Yices} & \textbf{Z3} & \textbf{ABC} \\
\hline
Pi          &   0.895 secs &    11.072 secs &    0.417 secs \\
Rho         &   4.509 secs &    15.776 secs &    5.879 secs \\
Theta       &  25.942 secs &    11.345 secs &   14.869 secs \\
Chi         &   5.059 secs &     3.930 secs &    1.327 secs \\
Iota        &  11.048 secs &     8.801 secs &  612.497 secs \\
KeccakF1600 & 745.206 secs & 15187.671 secs & 2910.549 secs \\
\hline
\end{tabular}
\end{table}

>>>>

\begin{compactitem}
\item Visual inspection and limited test vectors for (1) and (2)
\item Only provide proof certificates for message sizes of X to Y for digest sizes of \textbf{FIXME} for (3)
\item Visual inspection only for anything above Keccak for (3) in OpenSSLs C-code
\end{compactitem}

Effort (see above and then consider below):
\begin{compactitem}
 \item How long were we working on the Cryptol code?
 \item How long were we working on the SAW proofs?
 \item How long did it take to understand the OpenSSL code?
\end{compactitem}
We'll want to determine how long it took us and then try to infer how long that would have taken had we been working full time (40 hours/week).

Because of the modular design of the SHA-3 specification, the Cryptol implementation and OpenSSL implementation are programmatically more similar. 
The only strong discrepancy is that the OpenSSL implementation organizes the bits of the sponge construction. 
The affect of modular cryptographic design can be seen by the proof completing successfully in the Yices, Z3, and ABC SMT-solvers. 
Previous work on SHA2-256 was only successful with Yices \ref{cite:nfm-us}. Full results are show in table \ref{finaltable}.

\begin{table*}[b]
\caption{Proof Runtimes for SHA-3 with SMT-solvers}\label{finaltable}
\setlength{\tabcolsep}{13.5pt}
\begin{tabular}{|l|r|r|r|}
\hline
\textbf{Function} & \textbf{Yices} & \textbf{Z3} & \textbf{ABC} \\
\hline
Pi          &   0.895 secs &    11.072 secs &    0.417 secs \\
Rho         &   4.509 secs &    15.776 secs &    5.879 secs \\
Theta       &  25.942 secs &    11.345 secs &   14.869 secs \\
Chi         &   5.059 secs &     3.930 secs &    1.327 secs \\
Iota        &  11.048 secs &     8.801 secs &  612.497 secs \\
KeccakF1600 & 745.206 secs & 15187.671 secs & 2910.549 secs \\
\hline
\end{tabular}
\end{table*}
